Concurrent Programming with CXL

### 2020-06-21 book/paper.tex
Typo
Massaging

\section*{Exercises}
\begin{enumerate}
\item Run Figure~\ref{fig:cs} using CXL.  As there is no protection of the critical
section, mutual is violated, the assertion should fail, and a trace should be reported.
Now plug the following code \texttt{while True: pass;;}
just before entering the critical section
in Figure~\ref{fig:cs} and run CXL again.
Mutual exclusion is guaranteed but progress is violated.
CXL should print a short trace
to a state from which a terminating state cannot be reached.
Describe in English the difference in the failure reports before
and after inserting the code.
\item See if you can come up with some different approaches that satisfy both
mutual exclusion and progress.  Try them with CXL and see if they work or not.
If they don't, try to understand why.
Do not despair if you can't figure it out---as
we will find out, it is possible but not easy.
\end{enumerate}

@@

\begin{quote}
\begin{itemize}
\item Draw additional states and steps in Figure~\ref{fig:rdwr}
for three processes.
\end{itemize}
\end{quote}

A process that wants to read first waits until there are no writers:
\texttt{nwriters == 0}.  If so, it increments the number of readers.
Similarly, a
process that wants to write waits until there are no readers \emph{or} writers.
If so, it increments the number of writers.
The important invariants in this code are:
\begin{itemize}
\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} \ge n$,
\item $n$ writers at \texttt{@wcs} $\Rightarrow \mathtt{nwriters} \ge n$,
\item $(\mathtt{nreaders} \ge 0 \land \mathtt{nwriters} = 0) \lor
    (\mathtt{nreaders} = 0 \land 0 \le \mathtt{nwriters} \le 1)$.
\end{itemize}
It is easy to see that the invariants hold and imply the reader/writer
specification.
The solution also supports progress: if no process is in the critical
section then any process can enter.  Better still: if any reader is in the
critical section, any other reader is also able to enter.

@@

\glsadd{busy waiting}
While correct, it is not considered a good solution.
The solution is an example of what is called \emph{busy-waiting}
\index{busy waiting}
(aka spin-waiting):
\index{spin waiting}
processes spin in a loop until some desirable application-level condition is met.
Figure~\ref{fig:inc} also has an example of busy-waiting: the \texttt{main}
process waits for the other two processes to finish by checking their \texttt{done}
flags.
The astute reader might wonder if obtaining the mutex itself is an
example of busy-waiting.  After all, the CXL \texttt{synch} implementation
of \texttt{lock()} spins in a loop until the mutex is available
(see Figure~\ref{fig:spinlocks}).

However, there is a big difference.  As we pointed out, the processes that
are waiting for a spinlock are \emph{blocked}.
In most operating systems and programming language runtimes,
when a process acquires a lock, the process is placed on a scheduling
queue and stops using CPU cycles until the lock becomes available.
CXL supports this with the \texttt{synchS} module.
Busy waiting disables all this: even when using the \texttt{synchS}
module, readers and writers only get suspended temporarily in case
there is contention for the mutex.
A process trying to obtain a read or write lock in
Figure~\ref{fig:rwbusy} has to obtain the mutex repeatedly
until the read or write lock is available.

@@

\section*{Exercises}
\begin{enumerate}
\item Using busy waiting, implement a ``bound lock'' that allows
up to \texttt{M} processes to acquire it at the same time.  A bound lock
with \texttt{M == 1} is an ordinary lock.
You should define a constant \texttt{M} and two methods:
\texttt{acquire\_bound\_lock()}
and \texttt{release\_bound\_lock()}.
\item Write a test program for your bound lock
that checks that no more than \texttt{M} processes can acquire the
bound lock.
\end{enumerate}

@@

It is instructive to see what happens when a writer is in the critical
section and two readers try to enter.  The first reader successfully
acquires \texttt{rlock} and but blocks (!) when trying to acquire
\texttt{rwlock}, which is held by the writer.  The second reader blocks
trying acquire \texttt{rlock} because it is held by the first reader.
When the writer leaves the critical section, the first reader acquires
\texttt{rwlock}, sets \texttt{nreaders} to 1, and releases \texttt{rlock}.
\texttt{rwlock} is still held.
Then the second reader acquires \texttt{rlock} and, assuming the first
reader is still in the critical section, increments \texttt{nreaders} to 2
and enter the critical section \emph{without} acquiring \texttt{rwlock}.
\texttt{rwlock} is essentially jointly held by both readers.
It does not matter in which order they leave: the second will release
\texttt{rwlock}.

@@

\section*{Exercises}
\begin{enumerate}
\item Why do you suppose the code in Figure~\ref{fig:rwmulti} uses a flag
per process rather than a simple counter?
Can you write a version of this test that uses a counter instead
of a flag per process?
% \item Write a ``library implementation'' of reader/writer locks so multiple
% reader/writer locks can be instantiated.
\item Write a test program for ``bound locks'' (as implemented in the
exercises of the last chapter) that checks that up to \texttt{M} processes
can acquire the bound lock at the same time.
\end{enumerate}

@@

Imagine five philosopers sitting around a table, each with a plate of food in
front of them, and a fork between each two plates.  Each philosopher requires
two forks to eat.  To start eating, a philosopher first picks up the fork on
the left, then the fork on the right.  Each philospher likes to take breaks
from eating to think for a while.  To do so, the philosopher puts down both
forks.  The philosophers repeat this procedure.  Dijkstra had them repeating
this for ever, but for the purposes of this book, philosphers can leave
the table when they're not eating.

>>>>

\section*{Exercises}
\begin{enumerate}
\item Run Figure~\ref{fig:cs} using CXL.  As there is no protection of the critical
section, mutual exclusion is violated,
the assertion should fail, and a trace should be reported.
Now plug the following code \texttt{while True: pass;;}
just before entering the critical section
in Figure~\ref{fig:cs} and run CXL again.
Mutual exclusion is guaranteed but progress is violated.
CXL should print a short trace
to a state from which a terminating state cannot be reached.
Describe in English the difference in the failure reports before
and after inserting the code.
\item See if you can come up with some different approaches that satisfy both
mutual exclusion and progress.  Try them with CXL and see if they work or not.
If they don't, try to understand why.
Do not despair if you can't figure it out---as
we will find out, it is possible but not easy.
\end{enumerate}

@@

A process that wants to read first waits until there are no writers:
\texttt{nwriters == 0}.  Then it increments the number of readers.
Similarly, a
process that wants to write waits until there are no readers \emph{or} writers.
Then it increments the number of writers.
The important invariants in this code are:
\begin{itemize}
\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} \ge n$,
\item $n$ writers at \texttt{@wcs} $\Rightarrow \mathtt{nwriters} \ge n$,
\item $(\mathtt{nreaders} \ge 0 \land \mathtt{nwriters} = 0) \lor
    (\mathtt{nreaders} = 0 \land 0 \le \mathtt{nwriters} \le 1)$.
\end{itemize}
It is easy to see that the invariants hold and imply the reader/writer
specification.
The solution also supports progress: if no process is in the critical
section then any process can enter.  Better still: if any reader is in the
critical section, any other reader is also able to enter.

@@

\glsadd{busy waiting}
While correct, it is not considered a good solution.
The solution is an example of what is called \emph{busy-waiting}
\index{busy waiting}
(aka \emph{spin-waiting}):
\index{spin waiting}
processes spin in a loop until some desirable application-level condition is met.
Figure~\ref{fig:inc} also has an example of busy-waiting: the \texttt{main}
process waits for the other two processes to finish by checking their \texttt{done}
flags.
The astute reader might wonder if obtaining the mutex itself is an
example of busy-waiting.  After all, the CXL \texttt{synch} implementation
of \texttt{lock()} spins in a loop until the mutex is available
(see Figure~\ref{fig:spinlocks}).

However, there is a big difference.  As we pointed out, the processes that
are waiting for a spinlock are \emph{blocked}.
In most operating systems and programming language runtimes,
when a process acquires a mutex, the process is placed on a scheduling
queue and stops using CPU cycles until the mutex becomes available.
CXL supports this with the \texttt{synchS} module.
Busy waiting disables all this: even when using the \texttt{synchS}
module, readers and writers only get suspended temporarily in case
there is contention for the mutex.
A process trying to obtain a read or write lock in
Figure~\ref{fig:rwbusy} has to obtain the mutex repeatedly
until the read or write lock is available.

@@

\section*{Exercises}
\begin{enumerate}
\item Draw additional states and steps in Figure~\ref{fig:rdwr}
for three processes.
\item Using busy waiting, implement a ``bound lock'' that allows
up to \texttt{M} processes to acquire it at the same time.  A bound lock
with \texttt{M == 1} is an ordinary lock.
You should define a constant \texttt{M} and two methods:
\texttt{acquire\_bound\_lock()}
and \texttt{release\_bound\_lock()}.
\item Write a test program for your bound lock
that checks that no more than \texttt{M} processes can acquire the
bound lock.
\end{enumerate}

@@

It is instructive to see what happens when a writer is in the critical
section and two readers try to enter.  The first reader successfully
acquires \texttt{rlock} and but blocks when trying to acquire
\texttt{rwlock}, which is held by the writer.  The second reader blocks
trying acquire \texttt{rlock} because it is held by the first reader.
When the writer leaves the critical section, the first reader acquires
\texttt{rwlock}, sets \texttt{nreaders} to 1, and releases \texttt{rlock}.
\texttt{rwlock} is still held.
Then the second reader acquires \texttt{rlock} and, assuming the first
reader is still in the critical section, increments \texttt{nreaders} to 2
and enters the critical section \emph{without} acquiring \texttt{rwlock}.
\texttt{rwlock} is essentially jointly held by both readers.
It does not matter in which order they leave: the second will release
\texttt{rwlock}.

@@

\section*{Exercises}
\begin{enumerate}
\item Why do you suppose the code in Figure~\ref{fig:rwmulti} uses a flag
per process rather than a simple counter?
Can you write a version of this test that uses a counter instead
of a flag per process?
% \item Write a ``library implementation'' of reader/writer locks so multiple
% reader/writer locks can be instantiated.
\item Write a test program for ``bound locks'' (as implemented in the
exercises of the last chapter) that checks that up to \texttt{M} processes
can acquire the bound lock at the same time.
\item It can be useful for a process to \emph{downgrade} its write lock
to a read lock when its done writing but not done reading.  Implement
\texttt{downgrade}.  Note, you cannot do this simply by releasing the
write lock and then obtaining the read lock.
\end{enumerate}

@@

Imagine five philosopers sitting around a table, each with a plate of food in
front of them and a fork between each two plates.  Each philosopher requires
two forks to eat.  To start eating, a philosopher first picks up the fork on
the left, then the fork on the right.  Each philospher likes to take breaks
from eating to think for a while.  To do so, the philosopher puts down both
forks.  Each philosopher repeat this procedure.  Dijkstra had them repeating
this for ever, but for the purposes of this book, philosphers can leave
the table when they're not eating.
